New cardinality estimation algorithms for HyperLogLog sketches

### 2017-02-19 paper/paper.tex
fixed some typos

More practicable and already widely used for many applications is the HyperLogLog algorithm \cite{Flajolet2007} with a near-optimal space complexity $\symBigO(\symError^{-2} \log\log\symCardinality +\log \symCardinality)$. It has the nice property that partial results can be easily merged, which is a necessity for distributed environments. Unfortunately, the originally proposed estimation method has some problems to guarantee the same accuracy over the full cardinality range. Therefore, a couple of variants have been developed to correct the original estimate by empirical means \cite{Heule2013,Rhodes2015,Sanfilippo2014}.

@@

Sometimes not only the number of distinct elements but also a sample of them is needed in order to allow later filtering according to some predicate and estimating the cardinalities of corresponding subsets. In this case the k-minimum values algorithm \cite{Beyer2007} is the method of choice which also allows set manipulations like the construction of intersections, disjoint sets, or unions \cite{Dasgupta2015}. The latter operation is the only one that is natively supported by HyperLogLog sketches. A sketch that represents the set operation result is not always needed. One approach to estimate the corresponding result cardinality directly is based on the inclusion-exclusion principle, which however can become quite inaccurate, especially for small Jaccard indices of the input sets \cite{Dasgupta2015}. Alternatively, the HyperLogLog sketches can be combined with some additional data structure that allows estimating the Jaccard index. Together with the HyperLogLog cardinality estimates the error of the intersection size estimate can be reduced \cite{Pascoe2013, Cohen2016}, however, at the expense of a significantly larger memory footprint due to the additional data structure.

@@

\symProbability(\symRegValVariate \leq \symRegVal\vert\symPoissonRate)
\symProbability(\symRegValVariate \leq \symRegVal-1\vert\symPoissonRate)

>>>>

More practicable and already widely used in many applications is the HyperLogLog algorithm \cite{Flajolet2007} with a near-optimal space complexity $\symBigO(\symError^{-2} \log\log\symCardinality +\log \symCardinality)$. It has the nice property that partial results can be easily merged, which is a necessity for distributed environments. Unfortunately, the originally proposed estimation method has some problems to guarantee the same accuracy over the full cardinality range. Therefore, a couple of variants have been developed to correct the original estimate by empirical means \cite{Heule2013,Rhodes2015,Sanfilippo2014}.

@@

Sometimes not only the number of distinct elements but also a sample of them is needed in order to allow later filtering according to some predicate and estimating the cardinalities of corresponding subsets. In this case the k-minimum values algorithm \cite{Beyer2007} is the method of choice which also allows set manipulations like the construction of intersections, relative complements, or unions \cite{Dasgupta2015}. The latter operation is the only one that is natively supported by HyperLogLog sketches. A sketch that represents the set operation result is not always needed. One approach to estimate the corresponding result cardinality directly is based on the inclusion-exclusion principle, which however can become quite inaccurate, especially for small Jaccard indices of the input sets \cite{Dasgupta2015}. Alternatively, the HyperLogLog sketches can be combined with some additional data structure that allows estimating the Jaccard index. Together with the HyperLogLog cardinality estimates the error of the intersection size estimate can be reduced \cite{Pascoe2013, Cohen2016}, however, at the expense of a significantly larger memory footprint due to the additional data structure.

@@

\symProbability(\symRegValVariate' \leq \symRegVal\vert\symPoissonRate)
\symProbability(\symRegValVariate' \leq \symRegVal-1\vert\symPoissonRate)
